---
title: "US Air Pollution Report"
author: "Ivan Avila, Lance Santana, Cole Hammes, Gustavo Perez"
date: "12-11-24"
format: html
editor: visual
---

# Introduction

"Which areas in the United States have shown the best and worst air quality trends for ozone, carbon monoxide, sulfur dioxide, and nitrogen dioxide , and how have these pollutants changed over time in these regions?"

```{r}
library(tidyverse)
library(maps)
library(ggplot2)

airquality_data <- read_csv("pollution_2000_2023.csv")
```

# Lance Santana's Section

Reads and cleans pollution data set:

Analysis summary:

The initial analysis focused on extreme pollution conditions, particularly in California, due to its unexpected results, which contrasted with the high ozone levels typically expected near industrial hubs. This observation led to an investigation as to the potential effects of the quantity of ozone samples coming up as 0s.

Reads and cleans pollution data set:

```{r}
#Removes negative values from data
pollution_data <- airquality_data[airquality_data$`O3 Mean` > 0,]
```

Question #1a/b: Which cities have the highest & lowest recorded ozone pollution levels?

```{r}
#| code-fold: true
#Sorts highest & lowest 100 ozone samples into a data frame
top_oz_samples <- pollution_data[order(pollution_data$`O3 Mean`, decreasing = TRUE),]
top_oz_cities <- top_oz_samples[1:100, c("State", "City", "O3 Mean")]
top_oz_cities

bottom_oz_samples <- pollution_data[order(pollution_data$`O3 Mean`, decreasing = FALSE),]
bottom_oz_cities <- bottom_oz_samples[1:100, c("State", "City", "O3 Mean")]
bottom_oz_cities
```

Notably, data points surrounding San Francisco and Los Angeles were recorded among the lower than expected ozone levels.

Question #1c: Where are the highest and lowest recorded ozone levels in the United States?

```{r}
#| code-fold: true
#Creates us.cities, needed for mapping, contains city coordinators
data("us.cities")
us_cities <- us.cities
us_cities$name <- sapply(strsplit(us_cities$name," "), function(x) x[1])

#Set variable with highest ozone city coordinators
oz_cities_match_top_oz <- us_cities %>%
  filter(name %in% top_oz_cities$City)

#Set variable with lowest ozone city coordinators
oz_cities_match_bottom_oz <- us_cities %>%
  filter(name %in% bottom_oz_cities$City)

#Maps cities 
maps ::map("state")
points(oz_cities_match_top_oz$long, oz_cities_match_top_oz$lat, col = "red", pch = 7)
points(oz_cities_match_bottom_oz$long, oz_cities_match_bottom_oz$lat, col = "blue", pch = 7)
legend("bottomright", legend = c("Highest Ozone", "Lowest Ozone"), col = c("red", "blue"), pch = c(7, 7), title = "Legend")
title(main = "Cities by Highest and Lowest Ozone Pollution")
```

A calculation of samples with zero ozone pollution in California’s data set revealed that almost 20% of all zero ozone samples were from California. While this suggests limited ozone pollution, potential errors in data collection may contribute to these results.

Question #1d: What are some potential issues with the data that may result in skewed statistics?

```{r}
#| code-fold: true
#Potential Issues:
# 1. us.cities does not have all the cities listed in our data set

# 2. inaccurate sampling resulting in false 0 samples (evidenced by values less than 0)

# 3. data emphasis on highly populated areas resulting in a higher frequency of false sampling

#Counts the number of 0s in the data set
zero_count <- sum(airquality_data$`O3 Mean` == 0, na.rm = TRUE)
cat("There are ", zero_count, "instances of a 0 value for ozone.\n")

#Counts the number of negative values in the data set
neg_count <- sum(airquality_data$`O3 Mean` < 0, na.rm = TRUE)
cat("There are ", neg_count, "instances of a negative value for ozone.\n\n")

#Counts the number of 0s or negatives in Cali data
cali_count <- sum(airquality_data$State == "California" & airquality_data$`O3 Mean` <= 0, na.rm = TRUE)
cat("There are ", cali_count, "instances of a negative or 0 values for California")

cali_percent <- (cali_count / (zero_count + neg_count)) * 100
cat(" which makes up for ", cali_percent,"% of the total amount of 0 values in the O3 Mean dataset")
```

Question #2a:

What are the states that have the highest and lowest average ozone levels?

```{r}
#| code-fold: true
#Finds state average ozone
avg_oz_states <- data.frame(State = unique(pollution_data$State), Mean_03 = NA)
for (i in 1:nrow(avg_oz_states)){
  state_name <- avg_oz_states$State[i]
  state_info <- pollution_data[pollution_data$State == state_name, ]
  avg_oz_states$Mean_O3[i] <- mean(state_info$`O3 Mean`)
}

#Sorts by highest mean ozone
avg_oz_states <- avg_oz_states[order(avg_oz_states$Mean_O3),]

#Selects and prints highest three states from data
top_3_oz_states <- avg_oz_states[1:3, c("State", "Mean_O3")]
top_3_oz_states

#Sorts by lowest mean ozone
avg_oz_states <- avg_oz_states[order(-avg_oz_states$Mean_O3),]

#Selects and prints highest three states from data
bottom_3_oz_states <- avg_oz_states[1:3, c("State", "Mean_O3")]
bottom_3_oz_states
```

This was further evidenced by comparing the data disbursement of California and Tennessee, which has the highest average ozone. A comparison between the two bar graphs showed that California’s data is skewed to the right while still holding a lower mean than Tennessee.

Question #2b:

What does the distribution of ozone data look like for California compared to the highest and lowest ozone states?

```{r}
#| code-fold: true
#Collects data for Tennessee
tenne_data <- pollution_data %>%
  filter(State == "Tennessee")

#Collects data for Alaska
alask_data <- pollution_data %>%
  filter(State == "Alaska")

#Collects data for California
cali_data <- pollution_data %>%
  filter(State == "California")

#Graphs box plot from Tennessee data
ggplot(tenne_data, aes(x = `O3 Mean`)) +
  geom_boxplot(fill = "red", color = "black") + 
  labs(title = "Tennessee", x = "Ozone Samples") +
  xlim(-.0006, 0.102) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

#Graphs box plot from Alaska data
ggplot(alask_data, aes(x = `O3 Mean`)) +
  geom_boxplot(fill = "blue", color = "black") +
  xlim(-.0006, 0.102) +
  labs(title = "Alaska", x = "Ozone Samples")  +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

#Graphs box plot from California data
ggplot(cali_data, aes(x = `O3 Mean`)) +
  geom_boxplot(fill = "purple", color = "black") + 
  labs(title = "California", x = "Ozone Samples") +
  xlim(-.0006, 0.102) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
```

Question #3a:

What is the average ozone air pollution for California compared to the highest and lowest ozone states?

```{r}
#| code-fold: true
#Preparing mean state ozone data by grouping and ordering
oz_mean_by_state <- pollution_data %>%
  group_by(State) %>%
  summarize(average_ozone_mean = mean(`O3 Mean`, na.rm = TRUE)) %>%
  arrange(desc(average_ozone_mean))

#Calculate highest, lowest average state ozone and median average value
highest_ozone_state_value <- max(oz_mean_by_state$average_ozone_mean)
lowest_ozone_state_value <- min(oz_mean_by_state$average_ozone_mean)
median_ozone_state_value <- median(oz_mean_by_state$average_ozone_mean)

#Uses highest and lowest values to find highest and lowest states
higest_ozone_state <- oz_mean_by_state %>%
  filter(average_ozone_mean == highest_ozone_state_value)

lowest_ozone_state <- oz_mean_by_state %>%
  filter(average_ozone_mean == lowest_ozone_state_value)

#Prints median state ozone level
cat("Median State Ozone:", median(oz_mean_by_state$average_ozone_mean),"\n\n")

#Prints high state info 
cat("State with highest average ozone level:\n", higest_ozone_state$State, higest_ozone_state$average_ozone_mean,"\n\n") 

#Prints low state info 
cat("State with lowest average ozone level:\n", lowest_ozone_state$State, lowest_ozone_state$average_ozone_mean,"\n\n")

#Comparison of California to previous data
print("California Average Ozone:")
mean(cali_data$`O3 Mean`)
cat("California's average ozone is ", higest_ozone_state$average_ozone_mean - mean(cali_data$`O3 Mean`), " below the highest state average\n")
cat("California's average ozone is ", mean(cali_data$`O3 Mean`) - median(oz_mean_by_state$average_ozone_mean), " above the median\n\n")
```

Considering that other states may be misrepresented by the frequency of zero ozone samples, a bar graph was developed to compare states with zero and only non-zero values. The graph illustrates states with a higher frequency of zero-ozone tend to have a lower average ozone concentration.

Question #3b:

How do average ozone levels vary across states, and how do California, Tennessee, and Alaska compare to other states and the median in terms of ozone pollution?

```{r}
#| code-fold: true
#Adding color to select states
oz_mean_by_state$fill_color <- "gray"
oz_mean_by_state$fill_color[oz_mean_by_state$State == "California"] <- "purple"
oz_mean_by_state$fill_color[oz_mean_by_state$State == "Tennessee"] <- "red"
oz_mean_by_state$fill_color[oz_mean_by_state$State == "Alaska"] <- "blue"

#Graphing all states average ozone with dashed median line
ggplot(oz_mean_by_state, aes(x = reorder(State, average_ozone_mean), y = average_ozone_mean)) +
  geom_bar(stat = "identity", fill = oz_mean_by_state$fill_color) +
  geom_hline(yintercept = median(oz_mean_by_state$average_ozone_mean), color = "purple", linetype = "dashed", linewidth = 1) +
  scale_fill_identity() +
  labs(title = "State Mean Ozone Levels", x = "State", y = "Average State Ozone") +
  theme(
    axis.text.x = element_blank()
  )

#Printed summary
print("Blue is Alaska the minumum, red is Tennessee the maximum, and purple is California. The dashed line is the median state ozone value.")
```

Question #3c/d: Which states have the highest frequency of recorded zero values for ozone levels, and how do these states' average ozone levels compare to the average ozone levels across all states?

```{r}
#| code-fold: true
#Count the zero values recorded for each state
state_zero_count <- airquality_data %>%
  filter(`O3 Mean` == 0) %>%
  group_by(State) %>%
  summarise(zero_count = n()) %>%
  arrange(desc(zero_count))
state_zero_count

#Preparing mean state ozone data by grouping and ordering
oz_mean_by_state <- pollution_data %>%
  group_by(State) %>%
  summarize(average_ozone_mean = mean(`O3 Mean`, na.rm = TRUE)) %>%
  arrange(desc(average_ozone_mean))

#Adding color to select states puple for 10 and up, blue for < 10
oz_mean_by_state$fill_color <- "gray"
oz_mean_by_state$fill_color[oz_mean_by_state$State == "California"] <- "purple"
oz_mean_by_state$fill_color[oz_mean_by_state$State == "Pennsylvania"] <- "purple"
oz_mean_by_state$fill_color[oz_mean_by_state$State == "Colorado"] <- "purple"
oz_mean_by_state$fill_color[oz_mean_by_state$State == "Alaska"] <- "purple"
oz_mean_by_state$fill_color[oz_mean_by_state$State == "New York"] <- "purple"
oz_mean_by_state$fill_color[oz_mean_by_state$State == "Virginia"] <- "purple"
oz_mean_by_state$fill_color[oz_mean_by_state$State == "Oregon"] <- "blue"
oz_mean_by_state$fill_color[oz_mean_by_state$State == "Texas"] <- "blue"
oz_mean_by_state$fill_color[oz_mean_by_state$State == "Missouri"] <- "blue"
oz_mean_by_state$fill_color[oz_mean_by_state$State == "Massachusetts"] <- "blue"

#Graphing all states average ozone with dashed median line
ggplot(oz_mean_by_state, aes(x = reorder(State, average_ozone_mean), y = average_ozone_mean)) +
  geom_bar(stat = "identity", fill = oz_mean_by_state$fill_color) +
  geom_hline(yintercept = median(oz_mean_by_state$average_ozone_mean), color = "purple", linetype = "dashed", linewidth = 1) +
  scale_fill_identity() +
  labs(title = "State Mean Ozone Levels", x = "State", y = "Average State Ozone") +
  theme(
    axis.text.x = element_blank()
  )

#Printed summary
print("Purple represents the states with a zero count greater to or equal to 10, while blue states represent states with less than 10 but not zero.")
```

Conclusion:

Given the limits of this analysis, factors or potential correlations cannot be ruled out without additional analysis or consultation with outside information. With these caveats in mind, the effects of 0 ozone samples are uncertain and results show that Alaska is the best place for ozone pollution, Tennessee the worst, and California falls near average.

# Cole Hammes' Section

Filter the file to only include mainland US States for mapping analysis
```{r}
#| code-fold: true
filtered_states <- airquality_data %>%
  filter(!State %in% c("Alaska", "Hawaii", "District Of Columbia"))
```

Question 1: Over the last 23 years, which states have the highest levels of average daily sulfur pollution in the air?
```{r}
#| code-fold: true
sulfur_5 <- filtered_states %>%
  group_by(State) %>%
  summarize(average_sulfur = mean(`SO2 Mean`, na.rm = TRUE)) %>%
  arrange(desc(average_sulfur))%>%
  head(5)

print(sulfur_5)

```

```{r}
#| code-fold: true
us_states <- map_data("state") 
sulfur_5 <- sulfur_5 %>%
  mutate(state = tolower(State))  

sulfur_5_map <- us_states %>%
  inner_join(sulfur_5, by = c("region" = "state"))

ggplot() +
  geom_polygon(data = us_states, aes(x = long, y = lat, group = group),
               fill = "gray", color = "white") +  
  geom_polygon(data = sulfur_5_map, aes(x = long, y = lat, group = group, fill = average_sulfur),
               color = "white") +  
  scale_fill_gradient(low = "yellow", high = "red", name = "Avg Sulfur Level") +
  theme_minimal() +
  labs(title = "Top 5 States by Average Sulfur Levels", x = "Longitude", y = "Latitude")
```
After mapping out the states with the highest levels of sulfur dioxide pollution, we can see that New York, Pennsylvania, Kentucky, Wisconsin, and Virginia hold the highest spots. Besides New York, this makes a lot of sense because these states historically have economies that rely heavily on production and the burning of fossil fuels. In the case for New York, vast infrastruycture and a high population within New York were high contributors to sulfur pollution. 

Question 2: How has recent climate regulation shifted the burning of high sulfur content fossil fuels across the country?
```{r}
#| code-fold: true
filtered_states$Date <- as.numeric(substr(filtered_states$Date, 1, 4)) 
filtered_data <- subset(filtered_states, Date >= 2018 & Date <= 2023)

sulfur_5New <-filtered_data%>%
  group_by(State) %>%
  summarize(average_sulfur = mean(`SO2 Mean`, na.rm = TRUE)) %>%
  arrange(desc(average_sulfur))%>%
  head(5)
  
  print(sulfur_5New)
```
In this code chunk I analyzed the states with the highest daily sulfur levels over the last 5 years in the hopes of revealing the affects of new population sentiment towards coal production and the resulting shift in energy sources has affected daily sulfur dioxide levels. 

```{r}
#| code-fold: true
us_states <- map_data("state") 
sulfur_5New <- sulfur_5New %>%
  mutate(state = tolower(State))  

sulfur_5New_map <- us_states %>%
  inner_join(sulfur_5New, by = c("region" = "state"))

ggplot() +
  geom_polygon(data = us_states, aes(x = long, y = lat, group = group),
               fill = "gray", color = "white") +  
  geom_polygon(data = sulfur_5New_map, aes(x = long, y = lat, group = group, fill = average_sulfur),
               color = "white") +  
  scale_fill_gradient(low = "yellow", high = "red", name = "Avg Sulfur Level") +
  theme_minimal() +
  labs(title = "Top 5 States by Average Sulfur Levels", x = "Longitude", y = "Latitude")
```
As seen in the map, not only have the states with the highest levels shifted, but also, average daily sulfur dioxide levels decreased. This suggests that nationwide regulation has worked in recent years to limit, specifically, the burning of coal. The shift to renewable energy sources and natural gas has lowered the amount of sulfur in the air. 

```{r}
#| code-fold: true
top_5 <- c("New York", "Kentucky", "Pennsylvania", "Wisconsin", "Virginia")

top_cities <- airquality_data %>%
  filter(State %in% top_5) %>%           
  group_by(State, City) %>%                    
  summarize(city_sulf = mean(`SO2 Mean`),
            .groups = "drop") %>%
  arrange(State, desc(city_sulf)) %>%          
  group_by(State) %>%                          
  slice_max(city_sulf, n = 3) %>%             
  ungroup()                                   

print(top_cities)
```
This code chunk isolates the cities with the highest daily average sulfur levels to be used for later analysis. 

# Gustavo Perez's Section

1: Measure Carbon Monoxide in the country and pin point the cities with the highest and lowest rates.

```{r}
#| code-fold: true
highest_City<- airquality_data %>%
  filter(year(Date)==2023)%>%
  group_by(City)%>%
  summarize(M_CO=mean(`CO AQI`))%>%
  arrange(desc(M_CO))
head(highest_City,6)

low_City<- airquality_data %>%
  filter(year(Date)==2023)%>%
  group_by(City)%>%
  summarize(M_CO=mean(`CO AQI`))%>%
  arrange((M_CO))
head(low_City,3)
```

Filtering out cities with the highest and lowest air quality index of carbon monoxide, I was able to determine that cities such as Los Angeles, Phoenix, and Fresno have the highest levels of Carbon Monoxide in 2023. In contrast I was also able to determine the cities of Lompoc, Cheyenne, and Dallas have a significantly lower levels of Carbon Monoxide.

```{r}
#| code-fold: true
data("us.cities")
HI<-us.cities
LA<- us.cities%>%
  filter(name=='Los Angeles CA')
city_LA<- data.frame(x=LA$lat,y=LA$long)

LP<- us.cities%>%
  filter(name=='Lompoc CA')
city_LP<- data.frame(x=LP$lat,y=LP$long)

PH<- us.cities%>%
  filter(name=='Phoenix AZ')
city_PH<- data.frame(x=PH$lat,y=PH$long)

CH<- us.cities%>%
  filter(name== 'Cheyenne WY')
city_CH<- data.frame(x=CH$lat,y=CH$long)

DA<- us.cities%>%
  filter(name== 'Dallas TX')
city_DA<- data.frame(x=DA$lat,y=DA$long)

FR<- us.cities%>%
  filter(name== 'Fresno CA')
city_FR<- data.frame(x=FR$lat,y=FR$long)

map<-map_data('state')
ggplot()+
  geom_polygon(data=map,aes(x=long,y=lat,group=group),fill='white',color='black')+
  geom_point(data=city_LA,aes(x=y,y=x),color='red',size=3)+
  geom_text(data=city_LA,aes(x=y,y=x,label='Los Angeles'),vjust=1)+
          
  geom_point(data=city_FR,aes(x=y,y=x),color='red',size=3)+
  geom_text(data=city_FR,aes(x=y,y=x,label='Fresno'),vjust=-1)+
            
  geom_point(data=city_LP,aes(x=y,y=x),color='darkgreen',size=5,shape=18)+
  geom_text(data=city_LP,aes(x=y,y=x,label='Lompoc'),vjust=-1)+
            
  geom_point(data=city_PH,aes(x=y,y=x),color='red',size=3)+
  geom_text(data=city_PH,aes(x=y,y=x,label='Phoenix'),vjust=-1,size=3)+
            
  geom_point(data=city_CH,aes(x=y,y=x),color='darkgreen',size=5,shape=18)+
  geom_text(data=city_CH,aes(x=y,y=x,label='Cheyenne'),vjust=2,size=3)+  
    
  geom_point(data=city_DA,aes(x=y,y=x),color='darkgreen',size=5,shape=18)+
  geom_text(data=city_DA,aes(x=y,y=x,label='Dallas'),vjust=2,size=3)+
  labs(title='Highest and Lowest location of Carbon Monoxide in 2023',x='Long',y='Lat')
```

2: How has the average carbon monoxide (CO) level changed yearly in Phoenix, Los Angeles, and Lompoc?

```{r}
#| code-fold: true
find_city<-function(pollutant){
  
  chosen_City<- airquality_data %>%
    filter(year(Date)==2023)%>%
    group_by(City)%>%
    summarize(avg_pollutant=mean(.data[[pollutant]],na.rm = TRUE))%>%
    arrange(desc(avg_pollutant))
  head(chosen_City,3)
}

Pol_city<-function(x){
  CO_plot<-data.frame(
   index=integer(),
   CO_M=numeric())

  PH<-airquality_data %>%
    filter(City %in% c(x))%>%
    arrange((Date))


  for (i in 2000:2023){
    PH_data<- PH %>%
      filter(year(Date)==i)
    CO_data<-c(mean(PH_data$`CO AQI`,na.rm = TRUE))
    
    CO_plot <- rbind(CO_plot, data.frame(index = i, CO_M = CO_data))
  }
  
  
  ggplot (CO_plot,aes(x=index,y=CO_M))+
    geom_line(color='black',linewidth=1)+
    geom_point(color='red',size=2)+
    labs(title=paste('Average Carbon Monoxide Air Quality Index in',x,'from 2000-2023'),
        x="Year", y= 'Carbon Monoxide')
} 

Calexico_Pol<-Pol_city(c('Phoenix'))
Calexico_Pol

LA_Pol<-Pol_city(c('Los Angeles'))
LA_Pol

Lompoc_Pol<-Pol_city(c('Lompoc'))
Lompoc_Pol

```

After measuring how Carbon Monoxide has changed over 23 years, I found a similar trend of CO levels decreasing steadily. In the cities with the highest average, there is an exponential drop the first 5 years and has continued to drop steadily. Lompoc, having the lowest average, began already at a low average yet has still experienced an overall drop while having larger spikes of increased and decreased levels throughout the year. Regardless of what level they started, overall all locations are decreasing in Carbon Monoxide.

3.  How does Carbon Monoxide compare to other pollutants?

```{r}
#| code-fold: true
Pol_compare<-function(x){
  combine_p<-data.frame(
   index=integer(),
   Value=numeric(),
   Pollutant=character())
  
  
  
  PH<-airquality_data %>%
    filter(City %in% c(x))%>%
    arrange((Date))


  for (i in 2000:2023){
    PH_data<- PH %>%
      filter(year(Date)==i)
    CO_data<-c(mean(PH_data$`CO AQI`,na.rm = TRUE))
    NO2_data<-c(mean(PH_data$`NO2 AQI`,na.rm = TRUE))
    SO2_data<-c(mean(PH_data$`SO2 AQI`,na.rm = TRUE))
    O3_data<-c(mean(PH_data$`O3 AQI`,na.rm = TRUE))
    
    combine_p<-rbind(combine_p,data.frame(index=i,value=CO_data,Pollutant='Carbon Monoxide'))
    combine_p<-rbind(combine_p,data.frame(index=i,value=NO2_data,Pollutant='Nitrogen Dioxide'))
    combine_p<-rbind(combine_p,data.frame(index=i,value=SO2_data,Pollutant='Sulfur Dioxide'))
    combine_p<-rbind(combine_p,data.frame(index=i,value=O3_data,Pollutant='Ozone'))
  }
  
  
  ggplot (combine_p,aes(x=index,y=value,color=Pollutant,group=Pollutant))+
    geom_line()+
    geom_point()+
    labs(title=paste('Average Pollutants Air Quality Index in',x,'from 2000-2023'),
        x="Year", y= 'AQI Value')+
    scale_color_manual(values=c("Carbon Monoxide"='red',"Nitrogen Dioxide"='darkgrey',"Sulfur Dioxide"='black',"Ozone"='darkgreen'))
}

Pollutants_avg<-Pol_compare(c('Los Angeles'))
Pollutants_avg
```

After comparing Carbon Monoxide on a city level, it is similar to the rest of pollutant of decreasing steadily throughout the years. In Los Angeles, they're reaching a steady pace and Carbon Monoxide has the most similar trend with Sulfur Dioxide.

# Ivan Avila's Section

```{r}
#| code-fold: true
#Summary Stats for NO2
NO2_summary <- airquality_data %>%
  summarize(
    Mean_NO2 = mean(`NO2 AQI`, na.rm = TRUE),
    Median_NO2 = median(`NO2 AQI`, na.rm = TRUE),
    Min_NO2 = min(`NO2 AQI`, na.rm = TRUE),
    Max_NO2 = max(`NO2 AQI`, na.rm = TRUE),
    stdev_NO2 = sd(`NO2 AQI`, na.rm = TRUE),
    Observations = n()
  )

NO2_summary
```

```{r}
#| code-fold: true
#Identifying outliers function for respective state
identifying_outliers <- function(data, state_name) {
  state_data <- data %>%
    filter(State == state_name)
  
  stats <- state_data %>%
    summarise(
      Q1 = quantile(`NO2 AQI`, 0.25, na.rm = TRUE),
      Q3 = quantile(`NO2 AQI`, 0.75, na.rm = TRUE)
    ) %>%
    mutate(
      IQR = Q3 - Q1,
      lower_bound = Q1 - 1.5 * IQR,
      upper_bound = Q3 + 1.5 * IQR
    )
  
  outliers <- state_data %>%
    filter(`NO2 AQI` < stats$lower_bound | `NO2 AQI` > stats$upper_bound)
  return(outliers)
}
```

```{r}
#| code-fold: true
#Visualization function for outliers
visualize_outliers <- function(data, state_name) {
  state_data <- data %>% 
    filter(State == state_name) %>%
    mutate(Year = year(Date))
  
  outliers <- identifying_outliers(data, state_name) %>%
    mutate(Year = year(Date))
  
  outlier_plot <- ggplot(state_data, aes(x = factor(Year), y = `NO2 AQI`)) +
    geom_boxplot(alpha = 0.8)+
    geom_point(data = outliers, aes(x = factor(Year), y = `NO2 AQI`),
               color = 'blue', size = 1)+
    labs(
      title = paste("Outliers in NO2 AQI for", state_name, "by Year"),
      x = "Year",
      y = "NO2 AQI") +
    theme(axis.text.x = element_text(size = 7))
      
                    
  return(outlier_plot)
  
}
```

```{r}
#| code-fold: true
outliers_cali <- visualize_outliers(airquality_data, "California")

outliers_cali
```

Question: Which states have the highest average levels of nitrogen dioxide (NO2) AQI?

Task: Create a map highlighting states by their average NO2 AQI levels to visualize geographical patterns in pollution.

```{r}
#| code-fold: true
#Creates a United States Map for a respective year to compare NO2 values
filtered_year_2000 <- airquality_data %>%
  mutate(Year = year(Date)) %>%
  filter(Year == 2000)

#creates new dataset and specifically looks for states and reorder them
state_avg_NO2_2000 <- filtered_year_2000 %>%
  group_by(State) %>%
  summarize(Avg_NO2_AQI = mean(`NO2 AQI`, na.rm = TRUE)) %>%
  mutate(State = tolower(State))

#imports map
map_data_states <- map_data("state")

avg_NO2_merged_states_2000 <- map_data_states %>%
  left_join(state_avg_NO2_2000, by = c("region" = "State"))

state_center <- avg_NO2_merged_states_2000 %>%
  group_by(region) %>%
  summarize(
    long = mean(range(long)),
    lat = mean(range(lat)),
    Avg_NO2_AQI= unique(Avg_NO2_AQI)
  ) %>%
  filter(!is.na(Avg_NO2_AQI))

#creates map
ggplot(avg_NO2_merged_states_2000, aes(x = long, y = lat, group = group, fill = Avg_NO2_AQI)) +
  geom_polygon(color = "white") +
  geom_text(data = state_center, aes(x = long, y = lat, label = round(Avg_NO2_AQI, 1)),
            inherit.aes = FALSE, color = "black", size = 3) +
  scale_fill_gradient(low = "blue", high = "red", na.value = "grey") +
  labs(title = "Average NO2 AQI by State in 2000",
       fill = "Avg_NO2_AQI") 

```

```{r}
#| code-fold: true
#Creates a United States Map for a respective year to compare NO2 values
filtered_year_2023 <- airquality_data %>%
  mutate(Year = year(Date)) %>%
  filter(Year == 2023)

#creates new dataset and specifically looks for states and reorder them
state_avg_NO2_2023 <- filtered_year_2023 %>%
  group_by(State) %>%
  summarize(Avg_NO2_AQI = mean(`NO2 AQI`, na.rm = TRUE)) %>%
  mutate(State = tolower(State))

#imports map
map_data_states <- map_data("state")

avg_NO2_merged_states_2023 <- map_data_states %>%
  left_join(state_avg_NO2_2023, by = c("region" = "State"))

#Calculates the center of the states in order to plot numbers
state_center <- avg_NO2_merged_states_2023 %>%
  group_by(region) %>%
  summarize(
    long = mean(range(long)),
    lat = mean(range(lat)),
    Avg_NO2_AQI= unique(Avg_NO2_AQI)
  ) %>%
  filter(!is.na(Avg_NO2_AQI))

#creates map
ggplot(avg_NO2_merged_states_2023, aes(x = long, y = lat, group = group, fill = Avg_NO2_AQI)) +
  geom_polygon(color = "white") +
  geom_text(data = state_center, aes(x = long, y = lat, label = round(Avg_NO2_AQI, 1)),
            inherit.aes = FALSE, color = "black", size = 3) +
  scale_fill_gradient(low = "blue", high = "red", na.value = "grey") +
  labs(title = "Average NO2 AQI by State in 2023",
       fill = "Avg_NO2_AQI") 
```

```{r}
#| code-fold: true
#top 5 states within 2023
top_5_states_2023 <- state_avg_NO2_2023 %>%
  arrange(desc(Avg_NO2_AQI)) %>%
  head(5)

top_5_states_2023

#top 5 states within 2000
top_5_states_2000 <- state_avg_NO2_2000 %>%
  arrange(desc(Avg_NO2_AQI)) %>%
  head(5)

top_5_states_2000
```

Question: How have NO2 levels changed over time in a selected State? Task: Filter for a specific State and plot a line chart of NO2 Mean by year. Comparing Max Values of Pollutants

```{r}
#| code-fold: true
#Function filters and plots inputed city from dataset
analyze_state_pollution <- function(data, state_name) {
  state_data <- data %>%
    filter(State == state_name) %>%
    mutate(Year = as.integer(format(as.Date(Date, format = "%Y-%m-%d"), "%Y")))
  
#Checks if City input is indeed in the dataset  
  if (nrow(state_data) == 0) {
    stop(paste("No data available for", state_name))
  }
  
  state_yearly <- state_data %>%
    group_by(Year) %>%
    summarize(Mean_NO2 = mean(`NO2 AQI`, na.rm = TRUE),
              Max_NO2 = max(`NO2 AQI`, na.rm = TRUE),
              Max_O3 = max(`CO AQI`, na.rm = TRUE),
              Max_CO = max(`O3 AQI`, na.rm = TRUE),
              Max_SO2 = max(`SO2 AQI`, na.rm = TRUE))
  
  plot_mean_NO2 <- ggplot(state_yearly, aes(x = Year)) +
    geom_line(aes(y = Mean_NO2, color = "Mean NO2"), size = 1) +
    geom_point(aes(y = Mean_NO2, color = "Mean NO2"), size = 2) +
    labs(title = paste("Yearly Mean NO2 Levels in", state_name),
         x = "Year", y = "NO2 AQI",
         color = "Legend")
  
  plot_max_pollutants <- ggplot(state_yearly, aes(x = Year)) +
    geom_line(aes(y = Max_NO2, color = "Max NO2"), size = 1) +
    geom_line(aes(y = Max_O3, color = "Max O3"), size = 1, linetype = "dashed") +
    geom_line(aes(y = Max_CO, color = "Max CO"), size = 1, linetype = "dotted") +
    geom_line(aes(y = Max_SO2, color = "Max SO2"), size = 1, linetype = "dotdash") +
    labs(title = paste("Yearly Maximum Values of Pollutants in", state_name),
         x = "Year", y = "Pollutatnt AQI",
         color = "Pollutants") 
  return(list(Mean_NO2_Plot = plot_mean_NO2, MAx_Pollutants_Plot = plot_max_pollutants))
}
```

```{r}
#| code-fold: true
cali_plots <- analyze_state_pollution(data = airquality_data, state_name = "California")

cali_plots
```

```{r}
#| code-fold: true
col_plots <- analyze_state_pollution(data = airquality_data, state_name = "Colorado")

col_plots
```

# Project Conclusion
